### 2016-04-30 writing/latex/introduction.tex
tidy up a bit

\chapter{Introduction\label{cha:introduction}}
\section{What is a Permutation?}
In \emph{The Art of Computer Programming}\cite[p.~45]{Knuth:1997:ACP:260999} Donald Knuth defines
A \emph{permutation of \(n\) objects} is an arrangement of \(n\) distinct
objects in a row. When considering permutations we can consider them as
occuring on the set \(\nrange{n} = \setrange{1}{n}\), therefore a permutation
is a \emph{bijection} \(\pi:\nrange{n} \mapsto \nrange{n}\). We can write a
permutation \(\pi\) in two line notation
\begin{equation*}
\pi = \twoline{1,2,\dotso,n}{\pi(1),\pi(2),\dotso,\pi(n)}
\end{equation*}
However, the most frequent notation used in computer science is
\emph{one-line notation}, in this form we drop the top line of the two line
notation so are left with the following.
\begin{equation*}
\pi = \perm{\pi(1),\pi(2),\dotso,\pi(n)}
\end{equation*}

\begin{example}
There are 6 permutations on \(\nrange{3}\).
\begin{equation*}
\perm{1,2,3},\perm{1,3,2},\perm{2,1,3},\perm{2,3,1}, \perm{3,1,2}, \perm{3,2,1}
\end{equation*}
\end{example}

@@

The class of all permutations of length \(n\) is \(\mathfrak{S}_n\) and
the class has size \(n!\). The class of all permutations is
\(\mathfrak{S}=\bigcup_{i=0}^{\infty}\mathfrak{S}_i\).

\section{Classical Permutation Patterns}
Classical Permutation Patterns began to be studied as a result of Knuth's
statements about stack-sorting in \emph{The Art of Computer Programming}
\cite[p.~243, Ex.~5,6]{Knuth:1997:ACP:260999}.

\begin{definition}{Order isomorphism.}
    Two substrings \(\alpha_1\alpha_2\dotsm\alpha_n\) and
    \(\beta_1\beta_2\dotsm\beta_n\) are said to be \emph{order isomorphic}
    if they share the same relative order, \ie, \(\alpha_r<\alpha_s\) if and
    only if \(\beta_r<\beta_s\).
\end{definition}

A permutation \(\pi\) is said to \emph{contain} the permutation \(\sigma\) of
length \(k\) as a pattern (denoted \(\sigma \preceq \pi\)) if there is some
subsequence \(i_1i_2\dotsm{}i_n\) such that the sequence
\(\pi(i_1)\pi(i_2)\dotsm\pi(i_k)\) is order isomorphic to
\(\sigma(1)\sigma(2)\dotsm\sigma(k)\). If \(\pi\) does not contain
\(\sigma\), we say that \(\pi\) \emph{avoids} \(\sigma\).

For example the permutation \(\pi = \perm{2,4,1,5,3}\) contains the pattern
\(\sigma = \perm{2,3,1}\), since the second, fourth and fifth elements
(\(453\)) are order isomorphic to \(\perm{231}\).
This can be seen graphically below, the points order isomorphic to \(\sigma\)
are highlighted.
\begin{equation*}
    \pattern{2,4,5}{2,4,1,5,3}{}
\end{equation*}

@@

Knuth's statements were exercises in showing that the permutations avoiding the
pattern \(\perm{2,3,1}\) completely categorise permutations that are sortable to
the identity permutation using only a single stack, and that permutations avoiding the
pattern \(\perm{3,2,1}\) completely categorise permutations that are sortable to
the identity permutation using only a single queue with bypass.

\section{Mesh Patterns}
Mesh Patterns were introduced by \textcite{journals/combinatorics/BrandenC11} to capture explicit expansions
for certain permutation statistics. They are a natural extension of Classical
permutation patterns. A \emph{mesh-pattern} is a pair
\begin{equation*}
    p = (\tau,R)\text{ with } \tau \in \mathfrak{S}_k \text{ and } R \subseteq
    [0,k]\times [0,k].
\end{equation*}

The set \(R\) is called the \emph{mesh-set} of the mesh pattern \(p\).
The figure for a mesh pattern looks similar to that for a classical pattern with the
addition that we shade the unit square with bottom left corner \((i,j)\) for each \((i,j) \in R\):
\begin{equation*}
    \pattern{}{ 2, 1, 3 }{ 0/1, 0/2, 1/0, 1/1, 2/1, 2/2 }
\end{equation*}

We define containment, and avoidance, of the pattern \(p\) in the permutation
\(\pi\) on mesh patterns analogously to classical containment, and avoidance,
of \(\tau\) in \(\pi\) with the additional restrictions on the relative
position of the occurence of \(\tau\) in \(\pi\). These restrictions say that
no elements of \(\pi\) are allowed in the shaded regions of the figure.

\begin{example}
    The pattern \(p=\mperm{2,1,3}{\{(0,1),(0,2),(1,0),(1,1),(2,1),(2,2)\}}=
    \textpattern{}{ 2, 1, 3 }{ 0/1, 0/2, 1/0, 1/1, 2/1, 2/2 }\) is contained in
    \(\pi = \perm{3,4,2,1,5}\) but is not contained in \(\sigma = \perm{4,2,3,1,5}\).
\end{example}
\begin{proof}
    Let us consider the figure for the permutation \(\pi\) we only need to find one occurence.
    \begin{equation*}
        \pattern{1,3,5}{3,4,2,1,5}{0/2,0/3,0/4,
                                   1/0,1/1,1/2,
                                   2/0,2/1,2/2,
                                   3/2,3/3,3/4,
                                   4/2,4/3,4/4}
    \end{equation*}
    We have found an occurence of the pattern \(p\) in \(\pi\) and therefore
    \(\pi\) contains \(p\).

    Now we consider the figure for the permutation \(\sigma\). This permutation
    avoids \(p\) since for every occurence of the classical pattern \(\perm{2,1,3}\)
    there is at least one point in one of the shaded boxes. Consider the subsequence
    \(315\) in \(\sigma\), this is an occurence of \(\perm{2,1,3}\) but not the mesh
    pattern since the points with values \(4\) and \(2\) are in the shaded areas.
    This is shown in the figure below.

    \begin{equation*}
        \pattern{3,4,5}{4,2,3,1,5}{0/1,0/2,0/3,0/4,
                                   1/1,1/2,1/3,1/4,
                                   2/1,2/2,2/3,2/4,
                                   3/0,3/1,3/2,
                                   4/1,4/2,4/3,4/4}
    \end{equation*}
    This is true for all occurences of \(\perm{2,1,3}\) in \(\sigma\) and
    therefore \(\sigma\) avoids \(p\).
\end{proof}

We denote the avoidance sets for mesh patterns in the same way as for
classical patterns.

Given a mesh pattern \(p=(\sigma,R)\) we say that \(\sigma\) is the \emph{underlying
classical pattern} of \(p\).
\begin{note}
    \label{not:classmesh}
    Classical patterns are a subclass of mesh patterns the classical pattern \(\pi\)
    can be represented by a mesh pattern as \((\pi,\emptyset)\).
\end{note}

@@

Avoiding pairs of patterns of the same length with certain properties has
also been studied in the past, \textcite{MR2178749} considered avoiding a
pair of vincular patterns of length 3. \textcite{2015arXiv151203226B} study avoiding
a vincular and a covincular pattern simultaneously in order to achieve some
interesting counting results. However, very little work has been done on avoiding
a mesh pattern and a classical pattern simultaneously, in this work we aim to
establish some ground in this field by computing coincidences and Wilf-classes
and calculating some of the enumerations of avoiders of a mesh pattern of length
2 and a classical pattern of length 3.

>>>>

\chapter{Introduction\label{cha:introduction}}
\section{What is a Permutation?}
In \emph{The Art of Computer Programming} \cite[p.~45]{Knuth:1997:ACP:260999} Knuth states,
\enquote{A \emph{permutation of \(n\) objects} is an arrangement of \(n\) distinct
objects in a row}. When considering permutations we can consider them as
occuring on the set \(\nrange{n} = \setrange{1}{n}\), therefore a permutation
is a \emph{bijection} \(\pi:\nrange{n} \mapsto \nrange{n}\). A
permutation \(\pi\) can be written in two-line notation
\begin{equation*}
\pi = \twoline{1,2,\dotso,n}{\pi(1),\pi(2),\dotso,\pi(n)}
\end{equation*}
However, the most frequent notation used in computer science is
\emph{one-line notation}, in this form we drop the top line of the two line
notation and are left with the following.
\begin{equation*}
\pi = \perm{\pi(1),\pi(2),\dotso,\pi(n)}
\end{equation*}

\begin{example}
The 6 permutations on \(\nrange{3}\), in one-line notation, are
\begin{equation*}
\perm{1,2,3},\perm{1,3,2},\perm{2,1,3},\perm{2,3,1}, \perm{3,1,2}, \perm{3,2,1}
\end{equation*}
\end{example}

@@

The set of all permutations of length \(n\) is \(\mathfrak{S}_n\) and
has size \(n!\). The set of all permutations is
\(\mathfrak{S}=\bigcup_{i=0}^{\infty}\mathfrak{S}_i\). Note that \(\mathfrak{S}_0\)
has exactly one element, the empty permutation \(\varepsilon\). As a function
this is equivalent to the unique bijection \(\emptyset \mapsto \emptyset\), and
it's one-line representation is the empty string.

\section{Classical Permutation Patterns}
Classical permutation patterns began to be studied as a result of Knuth's
statements about stack-sorting in \emph{The Art of Computer Programming}
\cite[p.~243, Ex.~5,6]{Knuth:1997:ACP:260999}.

\begin{definition}{(Order isomorphism.)}
    Two substrings \(\alpha_1\alpha_2\dotsm\alpha_n\) and
    \(\beta_1\beta_2\dotsm\beta_n\) are said to be \emph{order isomorphic}
    if they share the same relative order, \ie, \(\alpha_r<\alpha_s\) if and
    only if \(\beta_r<\beta_s\).
\end{definition}

A permutation \(\pi\) is said to \emph{contain} the permutation \(\sigma\) of
length \(k\) as a \emph{pattern} (denoted \(\sigma \preceq \pi\)) if there is some
subsequence \(i_1i_2\dotsm{}i_k\) such that the sequence
\(\pi(i_1)\pi(i_2)\dotsm\pi(i_k)\) is order isomorphic to
\(\sigma(1)\sigma(2)\dotsm\sigma(k)\). If \(\pi\) does not contain
\(\sigma\), we say that \(\pi\) \emph{avoids} \(\sigma\).

For example the permutation \(\pi = \perm{2,4,1,5,3}\) contains the pattern
\(\sigma = \perm{2,3,1}\), since the second, fourth and fifth elements
(\(453\)) are order isomorphic to \(\perm{231}\).
This can be seen graphically below, the subsequence order isomorphic to \(\sigma\)
are highlighted.
\begin{equation*}
    \pattern{2,4,5}{2,4,1,5,3}{}
\end{equation*}

@@

Knuth's statements were exercises in showing that the permutations avoiding the
pattern \(\perm{2,3,1}\) are precisely the permutations that are sortable to
the identity permutation using a single stack, and that permutations avoiding the
pattern \(\perm{3,2,1}\) are precisely the permutations that are sortable to
the identity permutation using a single queue with bypass.

\section{Mesh Patterns}
Mesh patterns were introduced by \textcite{journals/combinatorics/BrandenC11} to
capture explicit expansions for certain permutation statistics. They are a
natural extension of classical permutation patterns. A \emph{mesh pattern} is a pair
\begin{equation*}
    p = (\tau,R)\text{ with } \tau \in \mathfrak{S}_k \text{ and } R \subseteq
    [0,k]\times [0,k].
\end{equation*}

The set \(R\) is called the \emph{mesh set} of the mesh pattern \(p\).
The figure for a mesh pattern looks similar to that of a classical pattern with the
addition that we shade the unit square with bottom left corner \((i,j)\) for each \((i,j) \in R\):
\begin{equation*}
    \pattern{}{ 2, 1, 3 }{ 0/1, 0/2, 1/0, 1/1, 2/1, 2/2 }
\end{equation*}

We define containment, and avoidance, of the pattern \(p\) in the permutation
\(\pi\) on mesh patterns analogously to classical containment, and avoidance,
of \(\tau\) in \(\pi\) with the additional restrictions on the relative
position of the occurence of \(\tau\) in \(\pi\). These restrictions say that
no elements of \(\pi\) are allowed in the regions of the figure corresponding
to shaded boxes in the mesh (see \cite[2]{journals/combinatorics/BrandenC11}
for the precise definition).

\begin{example}
    The pattern \(p=\mperm{2,1,3}{\{(0,1),(0,2),(1,0),(1,1),(2,1),(2,2)\}}=
    \textpattern{}{ 2, 1, 3 }{ 0/1, 0/2, 1/0, 1/1, 2/1, 2/2 }\) is contained in
    \(\pi = \perm{3,4,2,1,5}\) but is not contained in \(\sigma = \perm{4,2,3,1,5}\).
\end{example}

Let us consider the figure for the permutation \(\pi\). The subsequence \(325\) is
an occurrence of \(213\) in the classical sense and the remaining points of \(\pi\)
are not contained in the regions corresponding to the shaded boxes in \(p\).
\begin{equation*}
    \pattern{1,3,5}{3,4,2,1,5}{0/2,0/3,0/4,
                               1/0,1/1,1/2,
                               2/0,2/1,2/2,
                               3/2,3/3,3/4,
                               4/2,4/3,4/4}
\end{equation*}
The subsequence \(325\) is therefore an occurence of the pattern \(p\) in \(\pi\) and
\(\pi\) contains \(p\).

Now we consider the figure for the permutation \(\sigma\). This permutation
avoids \(p\) since for every occurence of the classical pattern \(\perm{2,1,3}\)
there is at least one point in one of the shaded boxes. For example, consider the
subsequence\(315\) in \(\sigma\), this is an occurence of \(\perm{2,1,3}\) but
not the mesh pattern since the points with values \(4\) and \(2\) are in the
regions corresponding to the boxes \((0,1)\) and \((0,2)\), which are shaded in
\(p\). This is shown in the figure below.

\begin{equation*}
    \pattern{3,4,5}{4,2,3,1,5}{0/1,0/2,0/3,0/4,
                               1/1,1/2,1/3,1/4,
                               2/1,2/2,2/3,2/4,
                               3/0,3/1,3/2,
                               4/1,4/2,4/3,4/4}
\end{equation*}
This is true for all occurences of \(\perm{2,1,3}\) in \(\sigma\) and
therefore \(\sigma\) avoids \(p\).

We denote the avoidance sets for mesh patterns in the same way as for
classical patterns. Given a mesh pattern \(p=(\sigma,R)\) we say that \(\sigma\) is the \emph{underlying
classical pattern} of \(p\).
\begin{note}
    \label{not:classmesh}
    Classical patterns can be thought of as a subset of mesh patterns: the
    classical pattern \(\pi\) can be represented by a mesh pattern as
    \((\pi,\emptyset)\).
\end{note}

@@

Avoiding pairs of patterns of the same length with certain properties has
also been studied in the past, \textcite{MR2178749} considered avoiding a
pair of vincular patterns of length 3. \textcite{2015arXiv151203226B} study avoiding
a vincular and a covincular pattern simultaneously in order to achieve some
interesting counting results. However, very little work has been done on avoiding
a mesh pattern and a classical pattern simultaneously. In this work we aim to
establish some ground in this field by computing coincidences and Wilf-classes
and calculating some of the enumerations of avoiders of a mesh pattern of length
2 and a classical pattern of length 3.
